Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-07-09 paper/submit-aas/referee-response2.txt

> However, the maps sent to the referee but not included in the paper do
> not ease my concerns on how robust their conclusions are. For example,
> the figures show that the assumption of a uniform density may be
> flawed given the number of features in the data. Especially the region
> around the spur shows a high density of large values of the unit
> weight error.
>
> Since their procedure starts with a selection of tentative members
> based on proper motions, it would be good if the authors could make a
> figure that shows both the stars selected according to the proper
> motion (i.e. the sample they already have) and the stars with 2
> parameter solutions that fall within the CMD selection. It would also
> be useful to assess the relative number of good and bad solutions in
> the blob and the spur, as well as in the regions around the gaps.
>
> The issue is not so much the quality of the data that is kept (the 5
> parameter solutions), but what does incompleteness in this dataset
> imply? The presence of stars with only 2 parameter solutions implies
> that such stars would not be selected according to the procedure
> followed by the authors, and their exclusion could potentially lead to
> "features". A plot as I propose would help to double check that the
> exclusion of such stars (if any) does not give rise to gaps or
> spurious over-densities.

TODO: Ana needs to crossmatch the 2par solutions with pan-starrs

> I would also like to ask the authors to overplot their selected GD-1
> stream stars (i.e. those shown in the bottom panel of Fig. 1), in the
> maps of the density of 2 parameter solutions, the unit weight
> error and the extinction map.

We have attached these figures to the response.

> Finally a small point regarding the presence of a progenitor. Given
> that it is very difficult to identify the feature the authors
> associated to the progenitor clearly from the figures presented, and
> its significance is not demonstrated in the paper, it does not really
> seem to deserve a mention in the abstract.

We have removed this statement from the abstract.

### 2018-07-19 paper/submit-aas/referee-response2.txt

> However, the maps sent to the referee but not included in the paper do
> not ease my concerns on how robust their conclusions are. For example,
> the figures show that the assumption of a uniform density may be
> flawed given the number of features in the data. Especially the region
> around the spur shows a high density of large values of the unit
> weight error.
>
> Since their procedure starts with a selection of tentative members
> based on proper motions, it would be good if the authors could make a
> figure that shows both the stars selected according to the proper
> motion (i.e. the sample they already have) and the stars with 2
> parameter solutions that fall within the CMD selection. It would also
> be useful to assess the relative number of good and bad solutions in
> the blob and the spur, as well as in the regions around the gaps.
>
> The issue is not so much the quality of the data that is kept (the 5
> parameter solutions), but what does incompleteness in this dataset
> imply? The presence of stars with only 2 parameter solutions implies
> that such stars would not be selected according to the procedure
> followed by the authors, and their exclusion could potentially lead to
> "features". A plot as I propose would help to double check that the
> exclusion of such stars (if any) does not give rise to gaps or
> spurious over-densities.

We have attached the requested figure to the report. We do not see any correlation between the off-track features described in this paper and the CMD-selected stars with 2-parameter solutions.

We have compared the fraction of good solutions in the spur, in the region between the stream and the spur, and in the stream at the same range of longitudes. We consider the longitude range -35 < phi1 < -30, and 3 ranges in phi2: (-0.4, 0.4), (0.4, 0.9), (0.9, 1.7) for the stream, region between stream and spur, and spur, respectively. In these 3 regions, we find 84%, 88%, 89% of the 5 parameter solution stars pass the unit weight error criteria.

@@

> I would also like to ask the authors to overplot their selected GD-1
> stream stars (i.e. those shown in the bottom panel of Fig. 1), in the
> maps of the density of 2 parameter solutions, the unit weight
> error and the extinction map.

We have attached these figures to the response. Again, while there is clearly structure in these maps, we do not see any relation between the data quality or incompleteness and the new off-track features reported in this paper. We do see issues with completeness at phi1 < -50 due to the scan pattern, and we discuss this in the text.

### 2018-07-23 paper/submit-aas/referee-response.txt

We thank the referee for a constructive report that has improved the quality of this work. Below, we have copied the remarks from the referee (indented by >) and have responded to each point in text that immediately follows. A few other minor changes have been made directly to the text and are indicated as bold text in the updated manuscript.

@@

The polyomial model (in phi1, phi2) is fit to the stream ridge points, then shifted up and down to define a selection region (as described in Section 3.1).

We agree that, to characterize the stream, it would be better to build a probabilistic model of the stream in all available data dimensions.
However, to be useful for future work, this would require simultaneously modeling the phase-space density of the stream so that the variations with distance and distribution of proper motions could be taken into account in a robust way.
As the focus of this work is mainly to point out unexpected, off-track features that appear related to the stream, we defer this more sophisticated modeling to future work.

=========================================
For the journal editor:

We have addressed each of the referee's comments in the attached response, but in doing so we have added a new figure that we believe improves the paper and directly addresses one of the referee's points. We realize that now the word count is a bit over the limit for ApJ letters (we count 4265 words). However, we do still feel that this result and paper are in the spirit of the Letters and would like to keep it in consideration for this journal. Please let us know if we should work on trimming the text, or else please advise. Thanks!

### 2018-07-23 paper/gd1.tex
doi for dataset

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Using data from the \gaia\ second data release combined with
\pans\ photometry, we present a sample of highly-probable members of the
longest cold stream in the Milky Way, GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects
high-contrast gaps along the stream, and (4) indicates
the existence of stream members perturbed off the main stream track.
These discoveries are only possible because of the exquisite astrometry
from \gaia, which permits a clean separation of the stream from Milky Way
stars.
The additional length and a proper treatment of the progenitor will aid in dynamical modeling of GD-1 for mapping the large-scale dark matter distribution.
The complex morphology of the stream points to a turbulent history; detailed phase-space properties of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.

@@

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible as an overdensity of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
\changes{At this location, between $\phi_1 \in (-18, -10)^\circ$, we find that the mean equatorial sky position of the stream stars is $(\alpha, \delta) = (177.01, 53.99)^\circ$ and the mean proper motion is $(\mu_\alpha \, \cos\delta, \mu_\delta) = (-7.78, -7.85) \pm 0.03~\masyr$, taking into account the covariance matrix for proper motions provided with the \gaia\ data.}

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Using data from the \gaia\ second data release combined with
\pans\ photometry, we present a sample of highly-probable members of the
longest cold stream in the Milky Way, GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) detects high-contrast gaps along the stream, and (3) indicates
the existence of stream members perturbed off the main stream track.
These discoveries are only possible because of the exquisite astrometry
from \gaia, which permits a clean separation of the stream from Milky Way
stars.
The additional length and a proper treatment of the progenitor will aid in dynamical modeling of GD-1 for mapping the large-scale dark matter distribution.
The complex morphology of the stream points to a turbulent history; detailed phase-space properties of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.

@@

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible as an overdensity of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
\changes{At this location, between $\phi_1 \in (-18, -10)^\circ$, we find that the mean equatorial sky position of the stream stars is $(\alpha, \delta) = (177.01, 53.99)^\circ$ and the mean proper motion is $(\mu_\alpha \, \cos\delta, \mu_\delta) = (-7.78, -7.85) \pm 0.03~\masyr$, taking into account the covariance matrix for proper motions provided with the \gaia\ data.}
\changes{The full data for the GD-1 region along with selection masks are available through Zenodo (\dataset[10.5281/zenodo.1295543]{https://doi.org/10.5281/zenodo.1295543})\footnote{See also \url{https://github.com/adrn/gd1-dr2}.}}
